### 2019-02-12 manuscript/manuscript.Rmd
a test

---
title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
output: 
  word_document:
    reference_docx: manuscript_template.docx   
author:
- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
- "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
- "^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston, RI 02881"
- "^*3*^ ORISE, Narragansett, RI 02882"
- "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
bibliography: manuscript.bib
csl: plos.csl
abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.  Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g. in urbanized northern Rhode Island).  Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis. This provides a robust and commonly scaled measurement to explore this data for long-term trends.  State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.  While yearly trends are useful, they do mask month-to-month variability differences across sites.  Additionally, while most sites track the yearly trend in decreasing water quality, there are bright spots with some sites improving over the 25 years. Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.  Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition. 
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(captioner)
source(here::here("R/functions.R"), encoding = "UTF-8")
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
```

```{r analysis, include=FALSE}
#got to figure out path stuff before render.
source(here::here("R/data_available.R"), encoding = "UTF-8") 
source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
source(here::here("R/figures.R"), encoding = "UTF-8")
```

# Introduction

1. Water quality in last several decades (Why Water Quality Paragraph(s))
  - Why of interest
    - cyanos stuff
    - CWA stuff
  - Understand long-term trends from local, to regional, to national
  - What variables are of interest and why
    - productivity
    - climate change
    - Nutrient enrichment stuff

2. Temporal and Spatial scaling of water quality trends Paragraph

3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
  - Shown to be stable in LAGOS (Maybe other studies in NE)
  - "Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds."

4. Rhode Island Paragraph (May be combined with #3)

4. Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.  First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.   In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends>.

# Methods

For this study we combined a long-term dataset on water quality of lakes in Rhode Island, with a trend analysis based on centered and scaled water quality values (i.e. z-scores) to find increasing or decreasing annual water quality trends.  Details are outlined below.

## Study Area and Data

The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).  The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 25 years. The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers. URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy. Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates. For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen. Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.

*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`*
![](../figures/ww_map.jpg)

In particular we selected data from the URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less. For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.  The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).  Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.

Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].  The authors found little change in water quality trends across this region [@Oliver_2017] .  We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].  Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a* from the LAGOSNE database.

## Trend Analysis

There are many different methods for analyzing time series data for trends.  One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.  For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds wer sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.  Thus, it is necessary to account for this type of site-specific variation.  This is similar to how the long-term temperature trends are analyzed using temperature anomolies [FILL IN GROUP DOING THIS].  The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.  This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the reference period.

### Z-score

The anomaly approach is very useful and is the approach we chose to follow.  However, using anomalies works best with a single measure (e.g temperature) or with mulitple measurements that are on the same scale.  The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.  For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees celsius to a high of 30 degrees celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.  To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.  This value is commonly referred to as a Z-score.  We use these z-scores to examine each water quality variable for a trend over the time period of 1990 to 2016.

Furthermore, since we are interest in water quality trends over time, we wanted to explore how each site was responding.  Thus, our z-scores were calculated over the reference period, 1990-2016, for each site.  In other words the long-term mean and standard deviation were calculated on a site-by-site basis.

### Kendall's Tau

Kendall's Tau and the Mann-Kendall test are commonly used to test for monotonic trends water quality data [@mann1945nonparametric; @helsel2002statistical; @hipel1994time].  We followed these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.

### Comparison of Rhode Island to the Region

Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].  The data we used in our analysis, the University of Rhode Island's Watershed Watch data, was inlcuded in that, but we were curious if regional trends were masking local scale trends in Rhode Island.  In addition to the analysis condcuted by Oliver et al. [-@Oliver_2017], we analyzed the same dataset but using the approach outlined here.

# Results

During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

*`r tab_num("ww_params")`*

```{r ww_params, echo=FALSE, results="asis"}
knitr::kable(ww_param_ranges, format = "markdown")
```

For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).

*`r tab_num("lagos_params")`*

```{r lagos_params, echo=FALSE, results="asis"}
knitr::kable(lagos_param_ranges, format = "markdown")
```

## State-wide trends in water quality

Test test test

Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

*`r fig_num(name = "temp_ri")`*

![](../figures/ww_temp_trends.jpg)

*`r fig_num(name = "chla_ri")`*

![](../figures/ww_chla_trends.jpg)

```{r no_early_tn, echo=FALSE}
ww_tn_kt_no9394 <- Kendall(ww_tn_df[c(-1,-2),]$year, ww_tn_df[c(-1,-2),]$mn_value)
```

Average yearly trends for nutrients showed weaker or no trends over time.  Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).  Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).  Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

*`r fig_num(name = "tn_ri")`*

![](../figures/ww_tn_trends.jpg)

*`r fig_num(name = "tp_ri")`*

![](../figures/ww_tp_trends.jpg)

## Regional trends in water quality

(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).

*`r fig_num(name = "tn_lagos")`*

![](../figures/lagos_tn_trends.jpg)

(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).

*`r fig_num(name = "tp_lagos")`*

![](../figures/lagos_tp_trends.jpg)

(`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).

*`r fig_num(name = "chla_lagos")`*

![](../figures/lagos_chla_trends.jpg)

# Discussion and conclusions

## Trends

### Local scale vs Regional/Global scale

### RI is downstream - Bryan will write

## Caveats

# Bibliography

### 2019-03-25 manuscript/manuscript.Rmd
Bryan's editorial suggestions and comments on the MS

@@
 title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
+author:
+- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
+  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A
+  ^*2*^
+- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
+  Ecology Division, Narragansett, RI 02882
+- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
+  RI 02881
+- ^*3*^ ORISE, Narragansett, RI 02882
+- *'^\*^  corresponding author: hollister.jeff@epa.gov'*
+bibliography: manuscript.bib
 output:
@@
     reference_docx: manuscript_template.docx
-author:
-- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
-- "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
-- "^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston, RI 02881"
-- "^*3*^ ORISE, Narragansett, RI 02882"
-- "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
-bibliography: manuscript.bib
+editor_options:
+  chunk_output_type: inline
 csl: plos.csl
-abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.
-Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g.
-in urbanized northern Rhode Island).
-Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis.
-This provides a robust and commonly scaled measurement to explore this data for long-term trends.
-State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen.
-Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.
-While yearly trends are useful, they do mask month-to-month variability differences across sites.
-Additionally, while most sites track the yearly trend in decreasing water quality, there are bright spots with some sites improving over the 25 years.
-Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.
-Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition.
-editor_options: 
-  chunk_output_type: console
+abstract: The University of Rhode Island's Watershed Watch Volunteer Monitoring
+  Program has been collecting water quality data on Rhode Island lakes and ponds for
+  close to three decades, allowing us to explore long-term trends in common water
+  quality parameters.  Not all lakes and ponds in the study area were sampled across
+  the full time period and lakes were often added in geographic clusters (e.g. in
+  urbanized northern Rhode Island).  Similar to the way long-term temperature records
+  are analyzed, we centered and scaled (i.e., the z-score) water quality measurements
+  on a per-station basis. This provides a robust and commonly scaled measurement to
+  explore this data for long-term trends.  State-wide aggregation shows increasing
+  temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus
+  is showing a decline, perhaps reflecting the management focus on phosphorus reductions.   While yearly trends are useful, they do mask month-to-month variability differences 
+  across sites.  Additionally, while most sites track the yearly trend in decreasing water
+  quality, there are bright spots with some sites improving over the 25 years. Contrary
+  to previously reported analyses that show relatively stable water quality at the
+  regional scale, our analysis shows that long-term water quality trends within Rhode
+  Island show some parameters improving (which ones?) while others are in decline (which ones?).  Importantly,
+  this analysis also points out the incredible value and importance of data from long-term
+  monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for
+  identifying trends in environmental condition.
 ---
@@
 ```{r setup, include=FALSE}
-library(knitr)
-library(captioner)
 source(here::here("R/functions.R"), encoding = "UTF-8")
@@
 fig_num <- captioner(prefix = "Figure")
+options(scipen = 2, digits = 5)
 ```
@@
 
-1. Water quality in last several decades (Why Water Quality Paragraph(s))
-  - Why of interest
-    - cyanos stuff
-    - CWA stuff
-  - Understand long-term trends from local, to regional, to national
-  - What variables are of interest and why
-    - productivity
-    - climate change
-    - Nutrient enrichment stuff
+Aquatic ecosystems have been altered as the result of human activities that modify bio-geochemical cycles on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global].
+Because of their position in the landscape, lakes integrate and serve as sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes].
+Increasing nutrient inputs, particularly of N (NOTE: I think we should spell out nitrogen and phosphorus throughout-WBM) and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
+Eutrophication has caused the frequency and severity of harmful algal blooms (HABs) to increase, and these blooms are predicted to worsen under a warming climate [@michalak2013record; @kosten2012warmer; @taranu2015acceleration].
+Increasing HABs will cause further declines in water quality, greater risks for human and animal health, and will be economically costly [@brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that concurrently reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
+These contrasting studies indicate that spatial differences and relationships between N, P, chlorophyll *a*, and HABs exist and that long-term studies are needed to identify trends at local, regional, and national scales.
 
-2. Temporal and Spatial scaling of water quality trends Paragraph
+Programs such as USEPA's National Lakes Assessment (NLA) provide data for continental scale water quality analysis.
+These analyses can be used for managing water resources by developing water quality criteria for N, P, chlorophyll *a*, and microcystin [@herlihy2013using; @yuan2014managing].
+Studying trends across broad spatial scales can elucidate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
+Continental scale data can also be used for water quality modeling across broad spatial scales including predicting lake trophic state and drawing inference on ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+These trophic state models indicate that landscape variables (i.e.
+ecoregion, elevation, and latitude) are important and that regional trends exist [ref???].
+Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
+Despite these challenges, it is important to study lake water quality at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
 
-3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
-  - Shown to be stable in LAGOS (Maybe other studies in NE)
-  - "Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds."
+Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Macroscale (i.e.
+subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
+This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
+Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
+Given what is known about long term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
 
-4. Rhode Island Paragraph (May be combined with #3)
+Examining long term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
+URI’s Watershed Watch is a XX year-old citizen science project that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
+Over the past XX years, volunteer monitors have been trained to collect *in situ* data and whole water samples during the growing season.
+These citizen science efforts allow for the collection of reliable data that can lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
+Here specifically, URIWW data contributed to not only the larger subcontinent extent study, but also allowed us to examine the long-term trends of Rhode Island.
 
-4.
 Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
 First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends>.
+In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOS project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [NEED CITATIONS].
 
@@
 
-The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
-The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 25 years.
-The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
-Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+The study area includes all lakes and ponds in the state of Rhode Island monitored by the University of Rhode Island's Watershed Watch (URIWW: pehaps we'd be better off with "watershed watch" instead of an acronym-wbm) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
+Watershed Watch is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 30 years.
+The program began in 1988 with the monitoring of 14 lakes.
+The program has now grown to include greater than 250 monitoring sites, 120 waterbodies (including rivers/streams, and estuaries), and 400 trained volunteers (past and present?).
+Watershed Watch is an integral part of Rhode Idland's state environmental data collection strategy and provides more than 90% of the state's multi-year baseline data on lakes.
+Data QA/QC is of paramount importance; volunteers receive both classroom and field training and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
 For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen.
@@
 
-*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`*
+*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites: NOTE (wbm): since all figures have two symbols /colors why not use the same throughout and make sure each figure has a legend")`*
 ![](../figures/ww_map.jpg)
 
-In particular we selected data from the URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
-Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.
+To understand the trends in lake temperature, total nitrogen, total phosphorus, and chlorophyll a we analyzed a subset of the watershed watch data.
+We chose sample sites with at least 10 years of data that were sampled at a depth of 2 m or less during the months of May to October from 1990 to 2016.
+The final dataset included `r nrow(ww_avail_data_locations)` lakes and ponds `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` of which had complete samples of all 4 parameters.
 
-Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+To place our results in context we reanalyzed the data from prior studies that looked at water quality trends across a larger region of the Northeastern and Midwestern United states that included 17 states (Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine) [@soranno2015building; @Oliver_2017].
 The authors found little change in water quality trends across this region [@Oliver_2017] .
@@
 There are many different methods for analyzing time series data for trends.
-One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.
+One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have spatial and temporal variation in sampling effort.
 For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds wer sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
 Thus, it is necessary to account for this type of site-specific variation.
-This is similar to how the long-term temperature trends are analyzed using temperature anomolies [FILL IN GROUP DOING THIS].
+This is similar to how the long-term temperature trends are analyzed using temperature anomolies [e.g., https://www.ncdc.noaa.gov/teleconnections/enso/indicators/sst/].
 The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.
 This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the reference period.
+
+NOTE(WBM): There is also local variation in history, lake morphometry, and ecology that can affect the data.
+Scaling and centering removes these confounding effects by converting all values to a mean of zero and a sd of 1.
 
@@
 In other words the long-term mean and standard deviation were calculated on a site-by-site basis.
+NOTE(wbm): So the index period will vary among lakes.
+I can image some reviewers will balk at this.
+Can we justify?
 
-### Kendall's Tau
+### Linear regression
 
-Kendall's Tau and the Mann-Kendall test are commonly used to test for monotonic trends water quality data [@mann1945nonparametric; @helsel2002statistical; @hipel1994time].
-We followed these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
+We used linear regression of z scores as a function of year to test for monotonic trends in the lake water quality data [@helsel2002statistical].
+Simply stated, this tests the null hypothesis that no trend exists (e.g.
+β1 = 0).
+If the null hypothesis is rejected a slope greater than zero indicates a positive or rising trend while a negative slope shows a negative or decreasing trend.
+Rejection of the null hypothesis for  lake temperature, chlorophyll *a*, total nitrogen, and / or total phosphorus indicates positive or negative trends for these parameters (depending on the sign of the slope).
+For each parameter we fit a regression line to the z scores as a function of year and tested the null hypothesis that no trend existed (e.g.
+$\beta$~1~ = 0).
 
@@
 
-Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-The data we used in our analysis, the University of Rhode Island's Watershed Watch data, was inlcuded in that, but we were curious if regional trends were masking local scale trends in Rhode Island.
-In addition to the analysis condcuted by Oliver et al.
-[-@Oliver_2017], we analyzed the same dataset but using the approach outlined here.
+Oliver et al.
+[@Oliver_2017] examined a larger dataset for the lakes of the Northeastern and Midwestern United States and found relatively stable water quality in the lakes over time.
+The University of Rhode Island's Watershed Watch data were inlcuded in this larger study.
+To assess whether regional trends were masking local scale trends in Rhode Island we re-analyzed the Oliver et al.
+[@Oliver_2017] dataset using the approach outlined above.
 
@@
 
-For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
 
@@
 
-Test test test
+Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
-Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-
-*`r fig_num(name = "temp_ri")`*
+*`r fig_num(name = "temp_ri")`* In this and subsequent figures values above or below the long term average are shown in red and blue respectively (NOTE: we should also use different symbols).
 
@@
 ```{r no_early_tn, echo=FALSE}
-ww_tn_kt_no9394 <- Kendall(ww_tn_df[c(-1,-2),]$year, ww_tn_df[c(-1,-2),]$mn_value)
+ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
+  tidy() %>%
+  slice(2) %>%
+  select(slope = estimate, p.value)
 ```
@@
 Average yearly trends for nutrients showed weaker or no trends over time.
-Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).
-Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Total nitrogen did have a significant (NOTE(wbm): Still, it looks like there could be a trend.
+Should we run the regression without the 1993 and 1994 data?) trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Since 1995, the yearly trend is not significant (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
+Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 
-(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
+In general, broad regional trends were less obvious and in the case of chlorophyll *a* and total phosphorus, the trends were not significant.
+In particular, chlorophyll *a* showed a slight, yet non-significant negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
+
+*`r fig_num(name = "chla_lagos")`*
+
+![](../figures/lagos_chla_trends.jpg)
+
+Total phosphorus shows a slight and non-significant increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`) and total nitrogen shows a slight negative non-significant trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
+
+*`r fig_num(name = "tp_lagos")`*
+
+![](../figures/lagos_tp_trends.jpg)
 
@@
 
-(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).
-
-*`r fig_num(name = "tp_lagos")`*
-
-![](../figures/lagos_tp_trends.jpg)
-
-(`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
-
-*`r fig_num(name = "chla_lagos")`*
-
-![](../figures/lagos_chla_trends.jpg)
-
 # Discussion and conclusions
 
+## Volunteer Monitoring Data
+
+### Why this is possible - Fantastic WW data
+
 ## Trends
+
+### Parameter by parameter discussion of trends
+
+### How do trends break down by trophic state?
 
@@
 
+## Management/Implications
+
+### Season Length Changes
+
 ## Caveats

### 2019-07-02 manuscript/manuscript.Rmd
Added citations

--- 2019-02-12 manuscript/manuscript.Rmd
@@
 ---
-title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
+title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
+author:
+- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
+  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
+  ^*2*^
+- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
+  Ecology Division, Narragansett, RI 02882
+- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
+  RI 02881
+- ^*3*^ ORISE, Narragansett, RI 02882
+- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
+bibliography: manuscript.bib
 output:
@@
     reference_docx: manuscript_template.docx
-author:
-- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
-- "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
-- "^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston, RI 02881"
-- "^*3*^ ORISE, Narragansett, RI 02882"
-- "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
-bibliography: manuscript.bib
-csl: plos.csl
-abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.
-Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g.
-in urbanized northern Rhode Island).
-Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis.
-This provides a robust and commonly scaled measurement to explore this data for long-term trends.
-State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen.
-Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.
-While yearly trends are useful, they do mask month-to-month variability differences across sites.
-Additionally, while most sites track the yearly trend in decreasing water quality, there are bright spots with some sites improving over the 25 years.
-Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.
-Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition.
 editor_options:
   chunk_output_type: console
+csl: ecology.csl
+abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
+Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine its effectiveness requires long-term datasets.
+A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States, however, trends at smaller scales may be masked.
+We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
+These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents.
+We assess for trends with z-scores (i.e.
+scaled anomalies) calculated on a per-station basis and examine yearly averages.
+Temperature and chlorophyll *a* are increasing.
+Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.
+Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.
+Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP all stable over time.
+In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
+Although not causal, this analysis suggests an association between lake temperature and productivity.
+Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.
+'
 ---
@@
 
-1. Water quality in last several decades (Why Water Quality Paragraph(s))
-  - Why of interest
-    - cyanos stuff
-    - CWA stuff
-  - Understand long-term trends from local, to regional, to national
-  - What variables are of interest and why
-    - productivity
-    - climate change
-    - Nutrient enrichment stuff
+Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global].
+Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes].
+Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
+This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
+These contrasting studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales.
 
-2. Temporal and Spatial scaling of water quality trends Paragraph
+Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
+These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
+Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
+Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+These trophic state models indicate that landscape variables (i.e.
+ecoregion, elevation, and latitude) are important and that regional trends exist.
+Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
+Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
 
-3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
-  - Shown to be stable in LAGOS (Maybe other studies in NE)
-  - "Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds."
+Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Macro-scale (i.e.
+subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
+This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
+Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
+Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.
 
-4. Rhode Island Paragraph (May be combined with #3)
+Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
+URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
+Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
+These efforts have been ongoing in some waterbodies since 1988.
+These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
+Additionally, the fact that volunteer monitors are used has allowed for this program to persist much longer than agency run programs that are often prone to uncertain funding.
+Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
+[@Oliver_2017], but also allowed us to examine the long-term trends in Rhode Island.
 
-4.
-Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
-First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends>.
+Thus, the goals of this study were to look at approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
+First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
+Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
+Another focus of this paper was to apply existing methodologies for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
+Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
 
@@
 
-The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
-The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 25 years.
+The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
+The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
 The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
+URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
 Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
-For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen.
-Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.
+For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
+Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory.
 
-*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`*
-![](../figures/ww_map.jpg)
+![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg)
 
-In particular we selected data from the URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
-Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.
+In particular we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+For this analysis we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
+For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
+Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
 
@@
 The authors found little change in water quality trends across this region [@Oliver_2017] .
-We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
-Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a* from the LAGOSNE database.
+We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
+Temperature data were not available, thus we only examine trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
 
-## Trend Analysis
+## Water Quality Trend Analysis
 
 There are many different methods for analyzing time series data for trends.
-One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.
-For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds wer sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
-Thus, it is necessary to account for this type of site-specific variation.
-This is similar to how the long-term temperature trends are analyzed using temperature anomolies [FILL IN GROUP DOING THIS].
-The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.
-This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the reference period.
+Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
+Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies (e.g., [@jones1996calculating]).
+The general approach is to calculate site-specific deviations from a long-term mean over a pre-determined reference period.
+This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change over the specified reference period.
 
@@
 
-The anomaly approach is very useful and is the approach we chose to follow.
-However, using anomalies works best with a single measure (e.g temperature) or with mulitple measurements that are on the same scale.
+Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees celsius to a high of 30 degrees celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
-To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.
-This value is commonly referred to as a Z-score.
-We use these z-scores to examine each water quality variable for a trend over the time period of 1990 to 2016.
+For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
+To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
+The resultant value is commonly referred to as a z-score.
+We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.
+Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.
+We refer to this approach as the site-specific z-scores.
 
-Furthermore, since we are interest in water quality trends over time, we wanted to explore how each site was responding.
-Thus, our z-scores were calculated over the reference period, 1990-2016, for each site.
-In other words the long-term mean and standard deviation were calculated on a site-by-site basis.
+### Summarizing site-specific z-scores
 
-### Kendall's Tau
+Methods for calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
+Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <fill in with zenodo link> [@zenodo_citation].
 
-Kendall's Tau and the Mann-Kendall test are commonly used to test for monotonic trends water quality data [@mann1945nonparametric; @helsel2002statistical; @hipel1994time].
-We followed these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
+![`r fig_num("z_schematic")`](../figures/z-score_schematic.jpg)
+
+The general steps, outlined in `r fig_num("z_schematic", display = "cite")` and listed below, are repeated for each of the water quality parameters.
+
+1.
+For each site, calculate the annual means, producing a single mean value for each site and year.
+This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].
+The per site means across years are assumed to be independent.
+2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
+3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
+4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
+
+### Linear regression on annual means
+
+Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
+We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
+For each parameter we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
+$\beta$~1~ = 0).
 
@@
 Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-The data we used in our analysis, the University of Rhode Island's Watershed Watch data, was inlcuded in that, but we were curious if regional trends were masking local scale trends in Rhode Island.
-In addition to the analysis condcuted by Oliver et al.
-[-@Oliver_2017], we analyzed the same dataset but using the approach outlined here.
+While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
+The analysis conducted by Oliver et al.
+[-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al.
+[-@Oliver_2017] but using the trend analysis approach outlined above.
 
@@
 
-During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
-
-*`r tab_num("ww_params")`*
+During the period of 1990 to 2016 Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 ```
+*`r tab_num("ww_params")`*
 
-For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
-
-*`r tab_num("lagos_params")`*
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
 
@@
 ```
+*`r tab_num("lagos_params")`*
 
@@
 
-Test test test
+Average annual scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
+Panel A.
+Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.
+Panel B.
+Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
 
-Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
 
-*`r fig_num(name = "temp_ri")`*
-
-![](../figures/ww_temp_trends.jpg)
-
-*`r fig_num(name = "chla_ri")`*
-
-![](../figures/ww_chla_trends.jpg)
+![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends.jpg)
 
 ```{r no_early_tn, echo=FALSE}
-ww_tn_kt_no9394 <- Kendall(ww_tn_df[c(-1,-2),]$year, ww_tn_df[c(-1,-2),]$mn_value)
+ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
+  tidy() %>%
+  slice(2) %>%
+  select(slope = estimate, p.value)
 ```
 
-Average yearly trends for nutrients showed weaker or no trends over time.
-Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).
-Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Mean annual trends for nutrients were weaker or showed no trend over time.
+The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score).
+Panel A.
+Yearly averaged TN z-scores from the URI Watershed Watch dataset.
+Panel B.
+Yearly averaged TN z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
+Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
+TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
+Panel A.
+Yearly averaged TP z-scores from the URI Watershed Watch dataset.
+Panel B.
+Yearly averaged TP z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
+The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).
+Panel A.
+Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.
+Panel B.
+Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
+Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].
 
-*`r fig_num(name = "tn_ri")`*
+![`r fig_num(name = "tn_trend")`](../figures/paired_tn_trends.jpg)
 
-![](../figures/ww_tn_trends.jpg)
+![`r fig_num(name = "tp_trend")`](../figures/paired_tp_trends.jpg)
 
-*`r fig_num(name = "tp_ri")`*
-
-![](../figures/ww_tp_trends.jpg)
+![`r fig_num(name = "np_trend")`](../figures/paired_np_trends.jpg)
 
@@
 
-(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
-
-*`r fig_num(name = "tn_lagos")`*
-
-![](../figures/lagos_tn_trends.jpg)
-
-(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).
-
-*`r fig_num(name = "tp_lagos")`*
-
-![](../figures/lagos_tp_trends.jpg)
-
-(`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
-
-*`r fig_num(name = "chla_lagos")`*
-
-![](../figures/lagos_chla_trends.jpg)
+In general, there was little to suggest broad regional changes.
+Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 
+Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
+neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
+One possible explanation for this is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
+Also, scale does indeed matter when trying to identify long-term water quality trends.
+Similar to the results of Oliver et al.
+[-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g.
+Northeastern and Mid-western United States).
+However, at the local scale of the state of Rhode Island there is a clear increasing trend in productivity.
+
 ## Trends
 
-### Local scale vs Regional/Global scale
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the TN:TP ratio are all relatively stable.
+These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., @).
+Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
+For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
+Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream].
+Thus differences between regional and state level trends are driven by complex processes.
 
-### RI is downstream - Bryan will write
+In addition to the annualized trends we address with this study, there are other trends that may be of interest.
+First, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
+Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
+Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
+Second, are trends influenced by the current state of a given waterbody?
+For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
+These questions are beyond the scope of this study, but all warrant further, careful investigation.
 
-## Caveats
+## Management Implications
+
+There are several broader management implications from the results of our analysis and of examining long-term water quality trends in general.
+In particular, this analysis provides much needed information about the long-term effects of current nutrient control efforts and identifies areas where additional information is required or a change in management approaches may be needed.
+First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term water quality trends.
+Specifically for this study, the results provide feedback to long time volunteer monitors, highlighting the importance of volunteer monitoring programs.
+Second, with information on long-term trends it is possible to adapt management approaches to address areas of concern.
+Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a longer growing season and associated higher productivity.
+
+There are several possible approaches.
+First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
+Green infrastructure approaches are one way to possibly achieve both goals [XXX].
+Additionally, within-lake approaches may also be used.
+A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species.
+Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
+
+## Data Analysis Approach
+
+The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is novel as we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
+This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].
+We simply built on these methods and adapted them for use with long-term water quality trends.
+And while other methods are valid and robust (e.g.
+[@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
+In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).
+Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.
+Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.
+Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
 

### 2019-07-05 manuscript/manuscript.Rmd
first edits

@@
 ---
-title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
+title: Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes
+  and Reservoirs
 author:
@@
 output:
+  html_document:
+    df_print: paged
   word_document:
@@
 csl: ecology.csl
-abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
-Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine its effectiveness requires long-term datasets.
-A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States, however, trends at smaller scales may be masked.
-We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
-URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
-These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents.
-We assess for trends with z-scores (i.e.
-scaled anomalies) calculated on a per-station basis and examine yearly averages.
-Temperature and chlorophyll *a* are increasing.
-Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.
-Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.
-Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP all stable over time.
-In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
-Although not causal, this analysis suggests an association between lake temperature and productivity.
-Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.
-'
+abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the
+  focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts
+  of eutrophication, but to determine the effectiveness requires long-term datasets.  A
+  recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE),
+  found stable water quality in the Northeast and Midwestern regions of the United
+  States; however, trends at smaller scales may be masked.  We address this by analyzing
+  the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW)
+  dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs
+  for over 25 years. These data, included in LAGOSNE, allow for comparison of water
+  quality trends at regional and state extents. We assess for trends with z-scores
+  (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature
+  and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing
+  trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total
+  z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll
+  is increasing, in spite of stable nutrient regimes. Although not causal, this analysis
+  suggests an association between lake temperature and productivity.  Additionally,
+  we demonstrate both the value of long-term monitoring programs, like URIWW, for
+  identifying trends in environmental condition, and the utility of site-specific
+  z-scores for analyzing for long-term trends. '
+  nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific
+  *a*, TN, TP, and TN:TP.  In short, productivity in Rhode Island lakes and reservoirs
 ---
@@
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These contrasting studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales.
+These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.
+Let's discuss it).
+Maybe...
+These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.(I think the edited sentence improves and clarifies the original)
 
+Attempts to explore these complex spatial dynamics are ongoing.
+(my attempt at intro sentence for paragraph).
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
-These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
-Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
-Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.
+Again, let's chat).
+Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
+Broad-scale data can also be used for water quality modeling across a range of spatial scales including predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
 These trophic state models indicate that landscape variables (i.e.
@@
 Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
+(thought- Maybe this paragraph just needs a introduction sentence.
+)
 
-Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper)(when used as adjectives, regional descriptors may or may not be capitalized depending on style (i.e.
+Chicago keeps lowercase) but should be capitalized when describing the region) United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
 Macro-scale (i.e.
@@
 This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the Northeastern and Midwestern US [@filstrup2014regional; @filstrup2018evidence].
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
@@
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
-Additionally, the fact that volunteer monitors are used has allowed for this program to persist much longer than agency run programs that are often prone to uncertain funding.
 Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
@@
 
-Thus, the goals of this study were to look at approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
-First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
-Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-Another focus of this paper was to apply existing methodologies for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
+Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
+First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature?
+Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states?
+Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
 Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
@@
 
-For this study we combined a long-term dataset on water quality of lakes in Rhode Island, with a trend analysis based on centered and scaled water quality values (i.e.
+For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e.
 z-scores) to find increasing or decreasing annual water quality trends.
@@
 The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
-The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
-Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+The program began in 1988, monitoring 14 lakes, and has now grown to include over 250 monitoring sites on over 120 waterbodies, including lakes, reservoir, rivers, streams, and estuaries with more than 400 trained volunteers.
+URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data (^I don't know what this means- is a common knowledge I'm missing?
+If so, ignore), and is an integral part of the state's environmental data collection strategy.
+Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
@@
 
-In particular we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
+In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
+For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, and `r length(total_p_10yr_sites)` samples for TP.
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
@@
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
-Temperature data were not available, thus we only examine trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
+Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
 
@@
 For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
-To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
+To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
 The resultant value is commonly referred to as a z-score.
@@
 We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
-For each parameter we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
+For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
 $\beta$~1~ = 0).
@@
 
-During the period of 1990 to 2016 Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the period of 1990 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 ![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
+
+- Reminder- need to add "a" to figure titles
 
@@
 
-In general, there was little to suggest broad regional changes.
+In general, there was little evidence to suggest broad regional changes.
 Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
@@
 neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
-One possible explanation for this is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
+One possible explanation is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
 Also, scale does indeed matter when trying to identify long-term water quality trends.
@@
 Northeastern and Mid-western United States).
-However, at the local scale of the state of Rhode Island there is a clear increasing trend in productivity.
+However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.
 
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the TN:TP ratio are all relatively stable.
-These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., @).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
+These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
-For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
-Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream].
-Thus differences between regional and state level trends are driven by complex processes.
+For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al.
+2002].
+Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007].
+Differences in regional and state level trends are driven by complex processes.
 
 In addition to the annualized trends we address with this study, there are other trends that may be of interest.
-First, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
+For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
 Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
-Second, are trends influenced by the current state of a given waterbody?
+Additionally, are trends influenced by the current state of a given waterbody?
 For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
@@
 Specifically for this study, the results provide feedback to long time volunteer monitors, highlighting the importance of volunteer monitoring programs.
-Second, with information on long-term trends it is possible to adapt management approaches to address areas of concern.
+Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
 Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a longer growing season and associated higher productivity.
@@
 First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
-Green infrastructure approaches are one way to possibly achieve both goals [XXX].
+Green infrastructure approaches are one way to possibly achieve both goals [see notes].
 Additionally, within-lake approaches may also be used.
 A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species.
-Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
+Some studies have shown reductions in both nutrients and algal biomass [see notes].
 
@@
 We simply built on these methods and adapted them for use with long-term water quality trends.
-And while other methods are valid and robust (e.g.
+While other methods are valid and robust (e.g.
 [@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.

### 2019-07-08 manuscript/manuscript.Rmd
minor edits, suggestions, and questions added

@@
 ---
-title: Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes
-  and Reservoirs
+title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
 author:
@@
 output:
-  html_document:
-    df_print: paged
   word_document:
@@
 csl: ecology.csl
-abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the
-  focus of lake management. Controlling phosphorus and nitrogen can mitigate the impacts
-  of eutrophication, but to determine the effectiveness requires long-term datasets.  A
-  recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE),
-  found stable water quality in the Northeast and Midwestern regions of the United
-  States; however, trends at smaller scales may be masked.  We address this by analyzing
-  the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW)
-  dataset.  URIWW has collected water quality data on Rhode Island lakes and reservoirs
-  for over 25 years. These data, included in LAGOSNE, allow for comparison of water
-  quality trends at regional and state extents. We assess for trends with z-scores
-  (i.e. scaled anomalies) calculated on a per-station basis and examine yearly averages.  Temperature
-  and chlorophyll *a* are increasing.  Total nitrogen (TN) shows a weak increasing
-  trend driven by low years in the early 1990s.  Total phosphorus (TP) and the total
-  z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll
-  is increasing, in spite of stable nutrient regimes. Although not causal, this analysis
-  suggests an association between lake temperature and productivity.  Additionally,
-  we demonstrate both the value of long-term monitoring programs, like URIWW, for
-  identifying trends in environmental condition, and the utility of site-specific
-  z-scores for analyzing for long-term trends. '
-  nitrogen:total phosphorus ratio (TN:TP) are stable.  Applying the site-specific
-  *a*, TN, TP, and TN:TP.  In short, productivity in Rhode Island lakes and reservoirs
+abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
+Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication but determining the effectiveness requires long-term datasets.
+A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE) found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at smaller scales may be masked.
+We address a smaller scale by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
+These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents.
+We assess for trends with z-scores (i.e.
+scaled anomalies) calculated on a per-station basis and examine yearly averages.
+Temperature and chlorophyll *a* are increasing.
+Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.
+Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.
+Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP.
+In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
+Although not causal, this analysis suggests an association between lake temperature and productivity.
+Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.
+'
 ---
@@
 Maybe...
-These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.(I think the edited sentence improves and clarifies the original)
+These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.
+(I agree with the previous comment and like the suggested sentence)
 
-Attempts to explore these complex spatial dynamics are ongoing.
-(my attempt at intro sentence for paragraph).
+Attempts to explore these spatial dynamics are ongoing (my attempt at intro sentence for paragraph).
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
@@
 Again, let's chat).
-Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
-Broad-scale data can also be used for water quality modeling across a range of spatial scales including predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
+Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
 These trophic state models indicate that landscape variables (i.e.
@@
 
-Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper)(when used as adjectives, regional descriptors may or may not be capitalized depending on style (i.e.
-Chicago keeps lowercase) but should be capitalized when describing the region) United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper)(It is my understanding that when using as a title, these should be capitalized, as compass directions not capitalized) United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
 Macro-scale (i.e.
@@
 First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature?
-Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states?
+Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United States?
 Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
@@
 The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
-The program began in 1988, monitoring 14 lakes, and has now grown to include over 250 monitoring sites on over 120 waterbodies, including lakes, reservoir, rivers, streams, and estuaries with more than 400 trained volunteers.
+The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
 URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data (^I don't know what this means- is a common knowledge I'm missing?
 If so, ignore), and is an integral part of the state's environmental data collection strategy.
-Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
 Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory.
+(I am not familiar with this certification.
+Is there a citation or certificate # to support?)
 
@@
 
-In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 (Throughout the methods and discussion this varies between 1990 and 1993.
+It would be good to be consistent) and 2016, 2) were sampled between May and October, 3) and were sampled at a depth of 2 meters or less.
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
 For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, and `r length(total_p_10yr_sites)` samples for TP.
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
@@
 Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
-The authors found little change in water quality trends across this region [@Oliver_2017] .
+The authors found little change in water quality trends across this region [@Oliver_2017] (This is repeated from the intro).
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
@@
 1.
-For each site, calculate the annual means, producing a single mean value for each site and year.
+For each site, calculate the annual means producing a single mean value for each site and year.
 This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].
@@
 $\beta$~1~ = 0).
+(This renders strangely)
 
@@
 
-Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
+Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].(This is repeated from the intro)  While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
 The analysis conducted by Oliver et al.
@@
 During the period of 1990 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+(add degrees to celsius?)
 
@@
 
-Average annual scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
+Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`), and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
 Panel A.
@@
 ```
-
+(Will the figures be reproduced in color for the print version? If not, should the captions be changed to not indicate colors?)
 Mean annual trends for nutrients were weaker or showed no trend over time.
@@
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
-TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
+Since 1995, the yearly trend shows a much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
+TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) (The slope and p-values are not the same as the figure) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
 Panel A.
@@
 In general, there was little evidence to suggest broad regional changes.
-Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio showed little change (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
-These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.(Is is worth discussing that the current nutrient load is stable but probably high due to previous loadings?)  These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
@@
 Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007].
-Differences in regional and state level trends are driven by complex processes.
+Thus, differences in regional and state level trends are driven by complex processes.
 
@@
 Additionally, are trends influenced by the current state of a given waterbody?
-For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
-These questions are beyond the scope of this study, but all warrant further, careful investigation.
+For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
+(Is this worth discussing further?
+There is at least one paper that could be mentioned.) These questions are beyond the scope of this study, but all warrant further, careful investigation.
 

### 2019-07-29 manuscript/manuscript.Rmd
added one sentence to trends section

@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.(Is is worth discussing that the current nutrient load is stable but probably high due to previous loadings?)  These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
+These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
+Although stable, there are likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
@@
 For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
-(Is this worth discussing further?
-There is at least one paper that could be mentioned.) These questions are beyond the scope of this study, but all warrant further, careful investigation.
+These questions are beyond the scope of this study, but all warrant further, careful investigation.
 
